A White Noise Approach to Evolutionary Ecology

### 2020-04-02 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
fig4Updated!

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued Markov processes. This approach allows us to compute the full implications of a stochastic process such as demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach can contribute to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of dynamical equations describing populations and communities which can then be used for rigorous investigations of the relationship between feedbacks of ecological and evolutionary processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document:
    toc: false
    number_sections: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
header-includes:
  - \usepackage{float}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'H')
```

@@

@@

To measure the total evolutionary change in species $i$ induced by species $j$, we form the metric $\Psi_{ij}=|\beta_{ij}|+|\gamma_{ij}|$. The top row of Figure \ref{net} displays interaction networks under weak and strong competition where the edge width connecting species $i$ and $j$ is proportional to $\alpha_{ij}\alpha_{ji}$. The bottom row of Figure \ref{net} displays the distributions of pairwise coevolutionary change, which we measure for species $i$ and $j$ via $\mathfrak{C}_{ij}=\Psi_{ij}\Psi_{ji}$, under weak and strong competition.

@@

@@

An alternative approach to breaking the assumption of normally distributed trait values is the development of explicit multilocus models. These models describe the contributions of alleles at particular loci in the genome to the development of quantitative traits. Tracking the fluctuations of allele frequencies then allows theoreticians to investigate the dynamics of phenotypic distributions that deviate from normality. This approach has a long history in theoretical quantitative genetics [@bulmer1980; @pmid7851785; @Kirkpatrick1727] and coevolutionary theory [@Nuismer2005; @Kopp2006; @Nuismer2007]. However, work to investigate the role of genetic architecture in mediating feedbacks between the dynamics of population abundances and the distributions of traits mediating ecological interactions has apparently only just begun [@Schreiber2018; @Patel2019].

@@

Using plant-pollinator communities as an example, the role of nitrogen mediating interspecific interactions has been reviewed by David, Storkey and Stevens [-@David2019] and the evolutionary ecology of the nutritional content of nectar has been reviewed by Parachnowitsch, Manson and Sletvold [-@Parachnowitsch2018]. These studies demonstrate the need for further research to understand how soil nutrient availability and organismal ecophysiology structures communities of plants and pollinators. Theoretical pursuits in this directions may profit from interfacing the framework we have outlined here with population-ecosystem models such as that developed by Fridley [-@Fridley2017].

@@

To understand patterns found in ecological communities, considerations must push beyond microevolutionary and contemporary ecological processes and consider the macroevolutionary dynamics of ancestral lineages leading to extant populations. Using sub-alpine flower communities as an example, one can observe a very strict ordering of phenology across broad geographic ranges. In particular, whether in the Colorado Rocky mountains (such as Gothic, Colorado) or on an outlier of the Idaho batholith (such as Kamiak butte near Palouse, Washington), one will almost surely observe a very conspicuous order of flowers emerging in early spring: at the very beginning of the season blooms _Claytonia lanceolata_ followed by _Erythronium grandiflorum_ which precedes _Delphinium nuttallianum_ (B. Week, personal observations). If contemporary phenological coevolution is rampant, why should this pattern be so well preserved across a thousand miles of rugged and diverse terrain? Although it would be exciting to find that these species repeatedly coevolved this pattern in each location, a more parsimoniuous hypothesis suggests the phenology and physiology of these species slowly evolved independently over macroevolutionary time scales to take advantage of the specific conditions available within each of these windows of the flowering season. However, this could not have carried out in the Rocky mountains since this terrain only became habitable just over ten thousand years ago as the glaciers of the Pleistocene began to retreat (Paul CaraDonna, personal communications). Hence, given these considerations, it appears that an understanding of early season phenology patterns should focus on how these communities are assembled as opposed to contemporary evolutionary dynamics. Indeed, recent work testing models of phylogeography ignores the potential for contemporary evolution and instead suggests alpine flower communities tend to follow neutral assembly where flowers merely compete for who can disperse to new habitat first, as opposed to a selective process where a regional species pool is filtered for those species adapted to the newly available habitat [@Marx2017].

>>>>

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued Markov processes. This approach allows us to compute the full implications of a stochastic process such as demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach can contribute to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of dynamical equations describing populations and communities which can then be used for rigorous investigations of the relationship between feedbacks of ecological and evolutionary processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document:
    extra_dependencies: ["float","flafter"]
    toc: false
    number_sections: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
header-includes:
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
#knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

@@

__Metric of pairwise coevolution__

@@

To measure the total evolutionary change in species $i$ induced by species $j$, we form the metric $\Psi_{ij}=|\beta_{ij}|+|\gamma_{ij}|$. Figure \ref{net} displays the joint distribution of $\mathfrak{C}_{ij}=\Psi_{ij}\Psi_{ji}$, our metric of pairwise coevolution, and the product of competition coefficients $\alpha_{ij}\alpha_{ji}$ for two simulated communities, both with richness $S=1000$. The solid contour represents the case of strong selection ($c=5\times10^{-7}$) and the dashed contour represents the case of weak sensitivity to competition ($c=1\times10^{-8}$). In both cases we see that $\mathfrak{C}_{ij}$ and $\alpha_{ij}\alpha_{ji}$ are, essentially, unrelated. However, this depiction is only representative of a specific set of parameters. Next, we provide analytical approximations of the covariance between selection gradients and competition coefficients and a numerical estimate for the relationship between pairwise coevolution and competition coefficients for a range of parameters.

@@

__Covariance of selection and competition as a function of diversity__

@@

Another approach to breaking the assumption of normally distributed trait values is the development of explicit multilocus models. These models describe the contributions of alleles at particular loci in the genome to the development of quantitative traits. Tracking the fluctuations of allele frequencies then allows theoreticians to investigate the dynamics of phenotypic distributions that deviate from normality. This approach has a long history in theoretical quantitative genetics [@bulmer1980; @pmid7851785; @Kirkpatrick1727] and coevolutionary theory [@Nuismer2005; @Kopp2006; @Nuismer2007]. However, work to investigate the role of complex genetic architectures in mediating feedbacks between the dynamics of population abundances and the distributions of traits mediating ecological interactions has apparently only just begun [@Patel2019].

@@

Using plant-pollinator communities as an example, the role of nitrogen mediating interspecific interactions has been reviewed by David, Storkey and Stevens [-@David2019] and the evolutionary ecology of the nutritional content of nectar has been reviewed by Parachnowitsch, Manson and Sletvold [-@Parachnowitsch2018]. These studies demonstrate the need for further research to understand how soil nutrient availability and organismal ecophysiology structures communities of plants and pollinators. Theoretical pursuits in this direction may profit from interfacing the framework we have outlined here with population-ecosystem models such as that developed by Fridley [-@Fridley2017].

@@

To understand patterns found in ecological communities, considerations must push beyond microevolutionary and contemporary ecological processes and consider the macroevolutionary dynamics of ancestral lineages leading to extant populations. Using sub-alpine flower communities as an example, one can observe a very strict ordering of phenology across a broad geographic range. In particular, whether in the Colorado Rocky mountains (such as Gothic, Colorado) or on an outlier of the Idaho batholith (such as Kamiak butte near Palouse, Washington), one will almost surely observe a very conspicuous order of flowers emerging in early spring: at the very beginning of the season blooms _Claytonia lanceolata_ followed by _Erythronium grandiflorum_ which precedes _Delphinium nuttallianum_ (B. Week, personal observations). If contemporary phenological coevolution is rampant, why should this pattern be so well preserved across a thousand miles of rugged and diverse terrain? Although it would be exciting to find that these species repeatedly coevolved this pattern in each location, a more parsimoniuous hypothesis suggests the phenology and physiology of these species slowly evolved independently over macroevolutionary time scales to take advantage of the specific conditions available within each of these windows of the flowering season. However, this could not have carried out in the Rocky mountains since this terrain only became habitable just over ten thousand years ago as the glaciers of the Pleistocene began to retreat (Paul CaraDonna, personal communications). Hence, given these considerations, it appears that an understanding of early season phenology patterns should focus on how these communities are assembled as opposed to contemporary evolutionary dynamics. Indeed, recent work testing models of phylogeography ignores the potential for contemporary evolution and instead suggests alpine flower communities tend to follow neutral assembly where flowers merely compete for who can disperse to new habitat first, as opposed to a selective process where a regional species pool is filtered for those species adapted to the newly available habitat [@Marx2017].
